Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-10-30 manuscript/manuscript.Rmd
edits to the manuscript

---
title             : "Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows"
shorttitle        : "A tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  Social Brain Sciences lab, Department of Humanities, Social and Political Sciences (D-GESS).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations form a cornerstone of experimental psychology, and have had a widerspread impact on our understanding of human cognition. However, the orthodox method for analysing such data -- comparing means between conditions -- is known to conceal valuable information about the timeline of psychological effects, such as their onset time and duration. The ability to reveal finer-grained, "temporal states" of cognitive processes can have important consequences for theory development by qualitatively changing the key inferences that are drawn from psychological data. Moreover, well-established analytical approaches, such as event history analysis, are able to evaluate the detailed shape of time-to-event distributions, and thus characterise the timeline of psychological states. One barrier to wider use of event history analysis, however, is that the analytical workflow is typically more time-consuming and complex than orthodox approaches. To help achieve broader uptake, in this paper we outline a set of tutorials that detail how to implement one distributional method known as discrete-time event history analysis. We illustrate how to wrangle raw data files and calculate descriptive statistics, as well as how to calculate inferential statistics via Bayesian and frequentist multi-level regression modelling. We touch upon several key aspects of the workflow, such as how to specify regression models, the implications for experimental design, as well as how to manage inter-individual differences. We finish the article by considering the benefits of the approach for understanding psychological states, as well as the limitations and future durections of this work. Finally, the project is written in R and freely available, which means the general approach can easily be adapted to other data sets, and all of the tutorials are avilable as .html files to widen access beyond R-users.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology"
wordcount         : "X"
bibliography      : ["extrareferences2.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf

@@

In experimental psychology, it is standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such mean-average comparisons have been the workhorse of experimental psychology over the last century, and have had a substantial impact on theory development and our understanding of the structure of cognition and brain function. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhatShapingRT2016; @panisWhenDoesInhibition2022; @wolkersdorferTemporalDynamicsSequential2020]. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].

As a simple illustration, Figure 1 shows the results of several simulated RT data sets, which show how mean-average comparisons between two conditions can conceal the shape of the underlying RT distributions. For instance, in examples 1-3, mean RT is always comparable between two conditions, while the distributios differ (Figure 1, top row). In contrast, in examples 4-6, mean RT is lower in condition 2 compared to condition 1, but the RT distributions differ in each case (Figure 1, bottom row). Therefore, a comparison of means would lead to a similar conclusion in examples 1-3, as well as examples 4-6, whereas a comparison of the distributions would lead to a different conclusion in every case.

@@

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the time course of psychological states. As such, the approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to discriminate between different theoretical accounts of psychological and/or brain-based processes.
For example, the distributions in Example 4 show that the effect starts around 200 ms and is gone by 600 ms. In contrast, in Example 5, the effect starts around 400 ms and is gone by 800 ms. And in the Example 6, the effect reverses around between 500 and 600 ms. What kind of theory or set of theories could account for such effects? Are there new auxiliary assumptions that theories need to adopt? And are there new experiments that need to be run to test the novel predictions that follow from these analyses? As we show later using concrete examples from past experimental data, for many psychological questions this "temporal states" information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes as well as adding a relatively under-used dimension -- the passage of time -- to our theory building toolkit.

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict time to death [e.g., @williammatthewmakehamLawMortalityConstruction1860].
The same logic has been applied to psychological time-to-event data, as previously demonstrated (Panis et al., 2020).
Here, in the paper, we hope to show the value of event history analysis for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience, as well as provide practical tutorials that provide step-by-step code and instructions in the hope that we can enable others to use event history analysis in a more routine, efficient and effective manner.

@@

In this paper, we focus on a distributional method for time-to-event data known as *discrete-time event history analysis*, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of event history analysis to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010], and our primary aim here is to introduce a set of tutorials, which explain **how** to do such analyses, rather than repeat in any detail **why** you should do them.

We then provide six different tutorials, each of which is written in the R programming language and publicly available on our Github and the Open Science Framework (OSF) pages, along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as a .html file, so that it can be viewed by any web browser, and thus available to those that do not use R.

In Tutorial 1a, we illustrate how to process or "wrangle" a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to *timed* accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the event history analysis of RT data for choice RT data. In Tutorial 3a, we illustrate how to fit the same type of multilevel regression model for RT data in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to event history analysis. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

@@

For a comprehensive background context to event history analysis, we recommend several excellent textbooks [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010]. Likewise, for general introduction to understanding regression equations, we recommend several introductory level textbooks (REFs). Our focus here is not on providing a detailed account of the underlying regression equations, since this topics has been comprehensively covered many times before. Instead, we want to provide an intuition to how event history analysis works in general as well as in the context of experimental psychology. As such, we only supply regression equations in the supplementary material (part D) and then refer to them in the text whenever relevant.

@@

To apply event history analysis (EHA), a.k.a. hazard analysis, one must be able to:

@@

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in the supplementary material (equations 1 and 2 in part A). The discrete-time hazard function gives you for each bin the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function gives you for each time bin the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins. As explained in part A of the supplementary material, the survivor function provides a context for interpreting the hazard function.

<!-- The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials contribute to each part of the hazard distribution. If a participant completes 100 trials in an experiment, and the survivor function reaches a probability of 0.03 at the end of timebin (400,500], then only 3% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in later parts of the hazard function would be wider and less precise compared to earlier parts.-->

@@

## 2.2 Event history analysis in the context of experimental psychology

### 2.2.1 A worked example

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a 2-button discrimination task, i.e., a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA can be extended with a discrete-time SAT analysis. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended analysis for choice RT data as EHA/SAT. As explained in part A of the supplementary material, the probability mass function provides a context for interpreting the conditional accuracy function.

Integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes. To illustrate, we consider a hypothetical choice RT example that is inspired by real data [@panisWhatShapingRT2016], but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target. Figure 3 shows that the early upswing in hazard is equal for both prime conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for short waiting times (< bin 6), responses always follow the prime (and not the target, as instructed). And then for longer waiting times, response hazard is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct. This is interesting because mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. And such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

Unlocking the temporal states of cognitive processes can be revealing in and of itself for theory development and the understanding of basic psychological processes. Possibly more importantly, however, is that it simultaneously opens the door to address many new and previously unanswered questions. Do all participants show similar temporal states or are there individual differences? Do such individual differences extend to those individuals that have been diagnosed with some form of psychopathology? How do temporal states relate to brain-based mechanisms that might be studied using other methods from cognitive neuroscience? And how much of theory in cognitive psychology would be in need of revision if mean-average comparisons were supplemented with a temporal states approach?

### 2.2.2 Implications for designing experiments

@@

We used `r my_r_citation$r` for all reported analyses. The content of the tutorials is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2018, @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

@@

@@

In general, it is important to visually inspect the functions first for each participant, in order to identify possible cheaters (e.g., a flat conditional accuracy function at .5 indicates (s)he was always guessing), outlying individuals, and/or different groups with qualitatively different behavior.

@@

When participants show qualitatively the same distributional patterns, one might consider to aggregate their data and make one plot (see Tutorial_1a.Rmd).

These distributional results suggest that the participant 6 is initially responding to the prime even though (s)he was instructed to only respond to the target, that response competition emerges in the incongruent prime condition around 300 ms, and that only slower responses are fully controlled by the target stimulus. Qualitatively similar results were obtained for the other five participants.

@@

To this end, Tutorial 1b illustrates how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of @panisWhatShapingRT2016, which includes two independent variables: mask type and prime type. As we use the same functional programming approach as in Tutorial 1a, we simply present the sample-based functions for participant 6 in Figure 5.

In the no-mask condition (column 1 in Figure 5), we observe a positive compatibility effect in the hazard and ca(t) functions, as congruent primes temporarily generate higher values for hazard and conditional accuracy compared to incongruent primes. However, when a (relevant, irrelevant, or lines) mask is present (columns 2-4), there is a negative compatibility effect in the hazard and conditional accuracy functions, as congruent primes temporarily generate *lower* values for hazard and conditional accuracy compared to incongruent primes.

@@

```{r plot-2ivs, fig.cap = "(ref:descr-plot5-caption)", out.width='95%'}
knitr::include_graphics("../Tutorial_1_descriptive_stats/figures/Plot_2IV_for_subject6_PanisSchmidt.png")
```

@@

Table 4 shows the summaries of the draws from the posterior distributions of the effects of congruent and incongurent primes relative to the blank prime condition in trials 500, 1000, and 1500, in terms of a point estimate (the mean) and the upper and lower bounds of the 95% highest posterior density interval. Also, by exponentiating the mean we obtain an effect size in terms of a hazard ratio.

@@

```{r int-table}
apa_table(
  int_tab,
  caption = "(ref:int-table-caption)",
  note = "(ref:int-table-note-caption)",
  placement = "H",
  font_size = "scriptsize",
  longtable = TRUE,
  repeat_header = TRUE,
  format.args = list(
       digits = c(0, 2,2,2,2,2,5),
        decimal.mark = ".", big.mark = "")
)
```

Based on Figure 6 and Table 4, we see that at the beginning of the experiment (trial 500), congruent and incongruent primes have a positive effect in time bin (200,240] on cloglog-hazard, relative to the cloglog-hazard estimate in the baseline condition (no prime; red striped lines in Figure 6). For example, the hazard ratio shows that the hazard of response occurrence for congruent primes is estimated to be 8.82 times higher than that for no-prime trials in bin (200,240] of trial 500. Incongruent primes also have a negative effect on cloglog-hazard in bins (320,360], (360,400], and (400,440]. For example, in bin (320,360], the hazard ratio shows that the hazard of response occurrence for incongruent prime is estimated to be .34 times smaller than that for no-prime trials. While the early positive effects reflect responses to the prime stimulus, the later negative effect for incongruent primes likely reflects response competition between the prime-triggered response (e.g., left) and the target-triggered response (e.g., right)

@@

These results show that the effect of prime-target congruency changes not only on the across-bin/within-trial time scale (variable period_9), but also on the across-trial/within-experiment time scale (variable trial_c). The fact that congruent primes generate negative effects for 80 ms (compared to no-prime trials) towards the end of the experiment, while incongruent primes generate negative effects for 120 ms throughout the experiment, strongly suggests the involvement of separate cognitive processes.

@@

Figure 8 shows one point (mean) and three highest posteriror density interval (50/80/95%) estimates for the effects of congruent and incongruent primes relative to neutral primes on logit-ca, for each time bin in trial numbers 500, 1000, and 1500.

@@

Table 5 shows the summaries of the draws from the posterior distributions of the effects of congruent and incongruent primes relative to the blank prime condition in trials 500, 1000, and 1500, in terms of a point estimate (the mean) and the upper and lower bounds of the 95% highest posterior density interval. Also, by exponentiating the mean we obtain an effect size in terms of an odds ratio.

@@

>>>>

---
title             : "Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows"
shorttitle        : "A tutorial on hazard analysis"
author: 
  - name          : "Sven Panis"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "ETH GLC, room G16.2, Gloriastrasse 37/39, 8006 Zürich"
    email         : "sven.panis@hest.ethz.ch"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : "Richard Ramsey"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "ETH Zürich"
authornote: |
  Neural Control of Movement lab, Department of Health Sciences and Technology (D-HEST).
  Social Brain Sciences lab, Department of Humanities, Social and Political Sciences (D-GESS).
  <!-- Enter author note here. -->
abstract: |
  Time-to-event data such as response times, saccade latencies, and fixation durations form a cornerstone of experimental psychology, and have had a widerspread impact on our understanding of human cognition. However, the orthodox method for analysing such data -- comparing means between conditions -- is known to conceal valuable information about the timeline of psychological effects, such as their onset time and duration. The ability to reveal finer-grained, "temporal states" of cognitive processes can have important consequences for theory development by qualitatively changing the key inferences that are drawn from psychological data. Moreover, well-established analytical approaches, such as event history analysis, are able to evaluate the detailed shape of time-to-event distributions, and thus characterise the timeline of psychological states. One barrier to wider use of event history analysis, however, is that the analytical workflow is typically more time-consuming and complex than orthodox approaches. To help achieve broader uptake, in this paper we outline a set of tutorials that detail how to implement one distributional method known as discrete-time event history analysis. We illustrate how to wrangle raw data files and calculate descriptive statistics, as well as how to calculate inferential statistics via Bayesian and frequentist multi-level regression modelling. Along the way, we touch upon several key aspects of the workflow, such as how to specify regression models, the implications for experimental design, as well as how to manage inter-individual differences. We finish the article by considering the benefits of the approach for understanding psychological states, as well as the limitations and future durections of this work. Finally, the project is written in R and freely available, which means the general approach can easily be adapted to other data sets, and all of the tutorials are avilable as .html files to widen access beyond R-users.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "response times, event history analysis, Bayesian multi-level regression models, experimental psychology, cognitive psychology"
wordcount         : "X"
bibliography      : ["extrareferences2.bib", "r-references.bib"]
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, donotrepeattitle"
appendix          : SupplementaryMaterial.Rmd
output            : papaja::apa6_pdf

@@

In experimental psychology, it is standard practice to analyse reaction times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such mean-average comparisons have been the workhorse of experimental psychology over the last century, and have had a substantial impact on theory development as well as our understanding of the structure of cognition and brain function. However, differences in mean RT conceal important pieces of information, such as when an experimental effect starts, how long it lasts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events [@panisHowCanWe2020; @panisNeuropsychologicalEvidenceTemporal2017; @panisStudyingDynamicsVisual2020; @panisTimecourseContingenciesPerceptual2009; @panisWhatShapingRT2016; @panisWhenDoesInhibition2022; @wolkersdorferTemporalDynamicsSequential2020]. Such information is useful not only for the interpretation of experimental effects under investigation, but also for cognitive psychophysiology and computational model selection [@panisAnalyzingResponseTimes2020].

As a simple illustration, Figure 1 shows the results of several simulated RT data sets, which show how mean-average comparisons between two conditions can conceal the shape of the underlying RT distributions. For instance, in examples 1-3, mean RT is always comparable between two conditions, while the distributions differ (Figure 1, top row). In contrast, in examples 4-6, mean RT is lower in condition 2 compared to condition 1, but the RT distributions differ in each case (Figure 1, bottom row). Therefore, a comparison of means would lead to a similar conclusion in examples 1-3, as well as examples 4-6, whereas a comparison of the distributions would lead to a different conclusion in every case.

@@

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the time course of psychological states. As such, the approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to discriminate between different theoretical accounts of psychological and/or brain-based processes.
For example, the distributions in Example 4 show that the effect starts around 200 ms and is gone by 600 ms. In contrast, in Example 5, the effect starts around 400 ms and is gone by 800 ms. And in the Example 6, the effect reverses between 500 and 600 ms. What kind of theory or theories could account for such effects? Are there new auxiliary assumptions that theories need to adopt? And are there new experiments that need to be performed to test the novel predictions that follow from these analyses? As we show later using published examples, for many psychological questions, such "temporal states" information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes, as well as adding a relatively under-used dimension -- the passage of time -- to the theory building toolkit.

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict the duration of time until people died [e.g., @williammatthewmakehamLawMortalityConstruction1860].
The same logic has been applied to psychological time-to-event data, as previously demonstrated (Panis et al., 2020).
Here, in the current paper, we hope to show the value of event history analysis for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience. Moreover, we provide tutorials that provide step-by-step code and instructions in the hope that we can enable others to use event history analysis in a more routine, efficient and effective manner.

@@

In this paper, we focus on a distributional method for time-to-event data known as *discrete-time event history analysis*, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of event history analysis to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010]. Indeed, our primary aim here is to introduce a set of tutorials, which explain **how** to do such analyses, rather than repeat in any detail **why** you may do them.

We provide six different tutorials, which are written in the R programming language and publicly available on our Github and the Open Science Framework (OSF) pages, along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as a .html file, so that it can be viewed by any web browser, and thus available to those that do not use R. Finally, the manuscript itself is written in R using the papaja() package [[insert REF for papaja() here]], which makes it computationally reproducible, in terms of the undeerlying data and figures.

In Tutorial 1a, we illustrate how to process or "wrangle" a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables.

In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to *timed* accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the event history analysis of RT data for choice RT data. In Tutorial 3a, we illustrate how to fit the same type of multilevel regression model for RT data in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to event history analysis. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

In tutorial 4, we illustrate one approach to planning how much data to collect in an experiment using EHA. We use data simulation techniques to vary sample size and trial count per condition until a certain degree of statistical power or precision is reached. [[more to come here, once we have written the tutorial]].

@@

For a comprehensive background context to event history analysis, we recommend several excellent textbooks [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010]. Likewise, for a general introduction to understanding regression equations, we recommend several excellent textbooks (Sven - See my comments in the notes file for relevant references here). Our focus here is not on providing a detailed account of the underlying regression equations, since this topic has been comprehensively covered many times before. Instead, we want to provide an intuition regarding how event history analysis works in general, as well as in the context of experimental psychology. As such, we only supply regression equations in supplementary material and then refer to them in the text whenever relevant.

@@

To apply event history analysis (EHA), one must be able to:

@@

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in the supplementary material (equations 1 and 2 in part A). The discrete-time hazard function gives you, for each time bin, the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function gives you for each time bin the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins.

The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials contribute to each part of the hazard distribution. If a participant completes 100 trials in an experiment, and the survivor function reaches a probability of 0.03 at the end of timebin (400,500], then only 3% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in later parts of the hazard function would typically be wider and less precise compared to earlier parts.

@@

## 2.2 Benefits of event history analysis

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. We do not cover these benefits in detail here, as these are more general topics that have been covered elsewhere in textbooks. Instead, we briefly summarise list the benefits below, and refer the reader to section F of Supplementary Materials for more detailed coverage of the benefits. A summary of the benefits are as follows:

1. Hazard functions are more diagnostic than density functions when one is interested in studying the detailed shape of a RT distribution (Holden et al., 2009).

2. RT distributions may differ from each other in multiple ways, and hazard functions allow one to capture these differences than mean-average comaprisons may conceal (Townsend, 1990).

3. EHA takes account of more of the data collected in a typical speeded response experiment, by virtue of not discarding right-censored observations. Trials with longer RTs are not discarded, but instead controbute to the risk set in each time bin.

4. Hazard modeling allows one to incorporate time-varying explanatory covariates, such as heart rate, electroencephalogram (EEG) signal amplitude, gaze location, etc. (Allison, 2010). This is useful for linking physiological effects to behavioral effects when performing cognitive psychophysiology [@meyerModernMentalChronometry1988].

5. EHA can help to solve the problem of model mimicry, i.e., the fact that different computational models can often predict the same mean RTs as observed in the empirical data, but not necessarily the detailed shapes of the empirical RT hazard distributions. As such, EHA can be a tool to help distinguish between competing theories of cognition and brain function.

## 2.3 Event history analysis in the context of experimental psychology

To make EHA more relevant to researchers studying cognitive psychology and cognitive neuroscience, in this section we provide a relevant worked example and consider implications that are relevant to that domain of research.

### 2.3.1 A worked example

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a 2-button discrimination task, i.e., a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA can be extended with a discrete-time SAT analysis. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function, which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended analysis for choice RT data as EHA/SAT.

Integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes. To illustrate, we consider a hypothetical choice RT example that is inspired by real data [@panisWhatShapingRT2016], but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target.

Figure 3 shows that the early upswing in hazard is equal for both prime conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for short waiting times (< bin 6), responses always follow the prime (and not the target, as instructed). And then for longer waiting times, the response hazard is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct.

This joint pattern of results is interesting because it can provide meaningfully different conclusions about psychological processes compared to conventional analyses, such as computing mean-average RT across trials. Mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. For instance, if mean-average RT was higher in incongruent than congruent trials, one may conclude that cognitive mechanisms that support interference control are working as expected across trials. But such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

Unlocking the temporal states of cognitive processes can be revealing for theory development and the understanding of basic psychological processes. Possibly more importantly, however, is that it simultaneously opens the door to address many new and previously unanswered questions. Do all participants show similar temporal states or are there individual differences? Do such individual differences extend to those individuals that have been diagnosed with some form of psychopathology? How do temporal states relate to brain-based mechanisms that might be studied using other methods from cognitive neuroscience? And how much of theory in cognitive psychology would be in need of revision if mean-average comparisons were supplemented with a temporal states approach?

### 2.3.2 Implications for designing experiments

@@

@@

## 3.3 Implementation

We used `r my_r_citation$r` for all reported analyses. The content of the tutorials, in terms of EHA and multi-level regression modelling, is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2018, @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

@@

In general, it is important to visually inspect the functions first for each participant, in order to identify individuals that may be guessing (e.g., a flat conditional accuracy function at .5 indicates that someone may be guessing), outlying individuals, and/or different groups with qualitatively different behavior.

@@

These distributional results suggest that the participant 6 is initially responding to the prime even though (s)he was instructed to only respond to the target, that response competition emerges in the incongruent prime condition around 300 ms, and that only slower responses are fully controlled by the target stimulus. Qualitatively similar results were obtained for the other five participants. When participants show qualitatively the same distributional patterns, one might consider to aggregate their data and make one plot (see Tutorial_1a.Rmd).

@@

To this end, Tutorial 1b illustrates how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of @panisWhatShapingRT2016, which includes two independent variables: mask type and prime type. As we use the same functional programming approach as in Tutorial 1a, we simply present the sample-based functions for participant 6 as part of Tutorial 1b for those that are interested.

@@

```{r plot-2ivs, fig.cap = "(ref:descr-plot5-caption)", out.width='95%'}
# knitr::include_graphics("../Tutorial_1_descriptive_stats/figures/Plot_2IV_for_subject6_PanisSchmidt.png")
```

@@

Table 4 shows the summaries of the draws from the posterior distributions of the effects of congruent and incongurent primes relative to the blank prime condition in trials 500, 1000, and 1500, in terms of a point estimate (the mean) and the upper and lower bounds of the 95% highest posterior density interval. Also, by exponentiating the mean we obtain an effect size in terms of a hazard ratio. For simplicity and ease of presentation, we only tabulate data for a subset of the design in the main text (trial 500 for congruent and incongruent conditions). For the full table, see Supplementary materials or Tutorual X.

@@

```{r int-table}
# apa_table(
#   int_tab,
#   caption = "(ref:int-table-caption)",
#   note = "(ref:int-table-note-caption)",
#   placement = "H",
#   font_size = "scriptsize",
#   longtable = TRUE,
#   repeat_header = TRUE,
#   format.args = list(
#        digits = c(0, 2,2,2,2,2,5),
#         decimal.mark = ".", big.mark = "")
# )
```

Based on Figure 6 and Supplementary Table XX, we see that at the beginning of the experiment (trial 500), congruent and incongruent primes have a positive effect in time bin (200,240] on cloglog-hazard, relative to the cloglog-hazard estimate in the baseline condition (no prime; red striped lines in Figure 6). For example, the hazard ratio shows that the hazard of response occurrence for congruent primes is estimated to be 8.82 times higher than that for no-prime trials in bin (200,240] of trial 500. Incongruent primes also have a negative effect on cloglog-hazard in bins (320,360], (360,400], and (400,440]. For example, in bin (320,360], the hazard ratio shows that the hazard of response occurrence for incongruent prime is estimated to be .34 times smaller than that for no-prime trials. While the early positive effects reflect responses to the prime stimulus, the later negative effect for incongruent primes likely reflects response competition between the prime-triggered response (e.g., left) and the target-triggered response (e.g., right)

@@

These results show that the effect of prime-target congruency changes not only on the across-bin/within-trial time scale (variable period_9), but also on the across-trial/within-experiment time scale (variable trial_c). The fact that congruent primes generate negative effects for 80 ms (compared to no-prime trials) towards the end of the experiment, while incongruent primes generate negative effects for 120 ms throughout the experiment, suggests the involvement of separate cognitive processes.

@@

Figure 8 shows one point (mean) and three highest posterior density interval (50/80/95%) estimates for the effects of congruent and incongruent primes relative to neutral primes on logit-ca, for each time bin in trial numbers 500, 1000, and 1500.

@@

Supplementary Table XX shows the summaries of the draws from the posterior distributions of the effects of congruent and incongruent primes relative to the blank prime condition in trials 500, 1000, and 1500, in terms of a point estimate (the mean) and the upper and lower bounds of the 95% highest posterior density interval. Also, by exponentiating the mean we obtain an effect size in terms of an odds ratio.

@@

## 4.4 Tutorial 4: Planning

This needs adding... RR to do this.
